Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-09-01 demystify.tex
updates

\chapter{MS Admission}
\chapterinfo{MS focuses on coursework and prepares you for industry, while PhD focuses on research and prepares you for academia or research labs. This section provides guidance on applying to MS programs.}

@@

In an MS application, you typically need to submit the similar materials as in a PhD application, e.g., transcripts, LoRs, SOP, and test scores.  However, research experience is not required, and LoRs can be from anyone who can speak about your academic or working abilities (e.g., professors, employers, etc). Moreover, your SOP you should focus on your academic and work (e.g., software development) background, why you want to get an MS, and how the program fits your goals. As with PhDs, GREs are often not required.  Moreover, some MS programs do not even require LoRs or SOPs.

\paragraph{Transcripts and Undergrad Background} Since research is not required, your undergrad background is more important in an MS application.  You should have a strong background in CS, e.g., through your undergrad degree.  If you do not have a CS degree, you might need to take some prerequisite courses before you can start the MS program.  GPA is also important, as it is often used as a filter for MS applications.

@@

\paragraph{LoRs} LoRs can be from professors, employers, or anyone who can speak about your academic or working abilities.  Unlike PhDs, your LoRs do not need to be from professors and talk about research experience. Many schools do not even require LoRs for MS.

\paragraph{Test Scores} GREs are often not required for MS applications.  However, some schools might still require them, and you should check with the school you're applying to.  If you have a low GPA, a high GRE score can help offset that.

>>>>

\chapter{MS Admission}
\chapterinfo{MS focuses on coursework and prepares you for industry, while PhD focuses on research and prepares you for academia or research.}

@@

You typically need to submit the similar materials as in a PhD application, e.g., transcripts, LoRs, SOP, and test scores.  However, research experience is not required, and LoRs can be from anyone who can speak about your academic or working abilities. Moreover, your SOP you should focus on your academic and work background, why you want to get an MS, and how the program fits your goals. As with PhDs, GREs are often not required. Some MS programs do not even require LoRs or SOPs.

\paragraph{Undergrad Background and GPA} Since research is not required, your undergrad background is more important in an MS application.  You should have a strong background in CS, e.g., through your undergrad degree.  GPA is also important, as it is often used as a filter for MS applications.

@@

\paragraph{LoRs} Unlike PhDs, your LoRs do not need to be from professors or talk about research experience. Many MS programs do not even require LoRs.

\paragraph{Test Scores} GREs are often not required for MS applications.  However, some schools might still require them, and you should check with the school you're applying to.  If you have a low GPA, a high GRE score can help offset that. As with PhD, English proficiency tests (TOEFL/IELTS) are required for international students.

\subsection{Funding}
Unlike PhD programs, which typically guarantee funding (\autoref{sec:funding}), MS programs are typically not funded. You have to pay tuition, which can be expensive. It is difficult for an MS student to get funding, e.g., through TA or RA positions, because these are typically reserved for PhD students.
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by the time they are, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.
\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. even have shortage of TA positions for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive, then a prof. might be willing to take you as an RA. Moreover, some schools do have TAs for MS students, and you can apply for these positions. For example, Mason CS dept has TAs for MS students (of course not for all of our 1000+ MS students, but quite a few); my courses in the past two years have MS TAs.
